Discrepancy, Coresets, and Sketches in Machine Learning

### 2019-03-29 komlos.tex
adds generalization result

\usepackage{amsmath}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{xcolor}
\usepackage{amsfonts}
\usepackage{ amssymb }
\usepackage{algorithm2e}
\newtheorem{fact}[theorem]{Fact}

%\newcommand{\zk}[1]{\textcolor{red}{ZK: #1}}
%\newcommand{\el}[1]{\textcolor{blue}{EL: #1}}
%\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}

@@

\begin{definition} [\bf Coreset Complexity] 
For a function $f:\X,\Q\rightarrow \R$ let $F(q) = \sum_{i=1}^{m} f(x_i,q)$ for a any set $\{x_1,\ldots,x_m\} \subset \X$.
For a set $S \subset [m]$ let $\tilde F(q) = \sum_{i \in S}w_i f(x_i,q)$ for some $w\in\R_+^m$ which is independent of $q$.
The coreset complexity of $f$ is the size of the smallest set $S$ such that $\forall q \in \Q \; |F(q)  - \tilde F(q)| \le \eps m$.
\end{definition}
\noindent The following facts are true for the common cases where $D_m = O(c/m)$ or $D_m = O(c/\sqrt{m})$. Although they were previously known (see e.g., \cite{phillips2009small}, Theorem 1.1) we give their proof here for completeness.
\begin{fact}\label{fct:eps1}
Any function $f$ with class discrepancy $D_m(f) = O(c/m)$ has coreset complexity of $O(c/\eps)$.
\end{fact}
\begin{fact}\label{fct:eps2}
Any function $f$ with class discrepancy $D_m(f) = O(c/\sqrt{m})$ has coreset complexity $O(c^2/\eps^2)$.
\end{fact}
\begin{proof}
\noindent  Consider the signed sum of error function $E(q) = \sum_{i=1}^{m} \sigma_i f(x_i,q)$ where $\sigma_i \in \{-1,1\}$.
Now consider $\tilde F_{+}(q) = F(q) + E(q)   = \sum_{i | \sigma_i=1} 2 f(x_i,q)$ and similarly $\tilde F_{-}(q) = F(q) - E(q)  =  \sum_{i | \sigma_i=-1} 2 f(x_i,q)$. We have that both $\tilde F_{+}(q)$ and $\tilde F_{-}(q)$ are approximations for $F(q)$ obtained by coresets of item weights of $2$ and an error at most $|\tilde F_{\pm}(q)- F(q)| =  |E(q)|$. The above is true for any choice of signs $\sigma$, specifically, for those minimizing $\max_q | E(q)|$.
By Definition we can select signs such that $|E(q)| \le m D_m(f)$.

Naturally, one could iterate this process.
At step one, create a coreset of size at most $n/2$ of items of weight $2$ and incur an error of at most $c$ or $c \sqrt{n}$.
Then, create a coreset of size $n/4$ of items of weight $4$. Here, you incur error of $2c$ or $2c\sqrt{n/2}  = c\sqrt{2n}$.
Note that the sum of errors is a geometric sequence that is asymptotically dominated by its last element. 
Halting the compression at $\Theta(c/\eps)$ or $\Theta(c^2/\eps^2)$ items respectively achieves the goal.
\end{proof}

Notice that these bounds are (asymptotically) tight in the context of unweighted coresets. Taking for example the bound of Fact~\ref{fct:eps1}, if we can guarantee the existence of an unweighted coreset of size $c/\eps$, then for $m$ items a coreset of size $m/2$ provides a sign assignment with an error of $\eps=2c/m$, leading to an upper bound of $O(c/m)$ on the class discrepancy.

>>>>

\usepackage{amsmath}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{xcolor}
\usepackage{amsfonts}
\usepackage{ amssymb }
\usepackage{algorithm2e}
\newtheorem{fact}[theorem]{Fact}
\usepackage{soul}

%\newcommand{\zk}[1]{\textcolor{red}{ZK: #1}}
\newcommand{\el}[1]{\textcolor{blue}{EL: #1}}
\newcommand{\add}[1]{\textcolor{blue}{ADD: #1}}
%\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}

@@

\begin{definition} [\bf Coreset Complexity] 
For a function $f:\X,\Q\rightarrow \R$ let $F(q) = \sum_{i=1}^{m} f(x_i,q)$ for a any set $\{x_1,\ldots,x_m\} \subset \X$.
For a set $S \subset [m]$ let $\tilde F(q) = \sum_{i \in S}w_i f(x_i,q)$ for some $w\in\R_+^m$ which is independent of $q$.
The coreset complexity of $f$ is the size of the smallest set $S$ such that $\forall q \in \Q \; |F(q)  - \tilde F(q)| \le \eps m$.
\end{definition}
\noindent The following facts are true for the common cases where $D_m = O(c/m)$ or $D_m = O(c/\sqrt{m})$. Although they were previously known (see e.g., \cite{phillips2009small}, Theorem 1.1) we give their proof here for completeness.
\begin{fact}\label{fct:eps1}
Any function $f$ with class discrepancy $D_m = O(c/m)$ has coreset complexity of $O(c/\eps)$.
\end{fact}
\begin{fact}\label{fct:eps2}
Any function $f$ with class discrepancy $D_m = O(c/\sqrt{m})$ has coreset complexity $O(c^2/\eps^2)$.
\end{fact}
\begin{proof}
\noindent  Consider the signed-sum error function $E(q) = \sum_{i=1}^{m} \sigma_i f(x_i,q)$ where $\sigma_i \in \{-1,1\}$.
Now consider $\tilde F_{+}(q) = F(q) + E(q)   = \sum_{i | \sigma_i=1} 2 f(x_i,q)$ and similarly $\tilde F_{-}(q) = F(q) - E(q)  =  \sum_{i | \sigma_i=-1} 2 f(x_i,q)$. 
We have that both $\tilde F_{+}(q)$ and $\tilde F_{-}(q)$ are approximations for $F(q)$ obtained by coresets of item-weight $2$. 
The error is at most $|\tilde F_{\pm}(q)- F(q)| =  |E(q)|$. 
The above is true for any choice of signs $\sigma$, specifically, for those minimizing $\max_q | E(q)|$.
By Definition we can select signs such that $|E(q)| \le m D_m$.

\add{
Naturally, one could iterate this process. Starting with $n$ items and ending with $m$.
Let $F_t$ denote the (unweighted) sum of functions $f$ after $t$ iterations and $n_t$ denote the cardinality of the coreset. 
\[
F = F_0 = 2F_1 \pm n_0D_{n_0} = 4F_{2} \pm 2n_1D_{n_1} \pm n_0D_{n_0} = \ldots = 2^T F_T \pm \sum_{t=0}^{T-1} 2^t n_t D_{n_t}
\]
Here $T$ stand for the total number of iterations needed for reaching a coreset of size $m$. 
Given $n_t \le n/2^{t}$ and the polynomial dependence of $D_m$ on $m$ we have 
$\sum_{t=0}^{T-1} 2^t n_t D_{n_t} \le n \sum_{t=0}^{T-1} D_{n/2^t} = n\cdot O(D_{n_T})$ for all $T$.
Specifically, for $T$ such that $n_T \approx m$, setting the weights $w = 2^T$ we have $F(q) = \tilde F(q) \pm O(nD_m)$.
Substituting the above expressions for $D_m$ completes the proof.
}
\st{
Naturally, one could iterate this process.
Noting that the last term in above is a geometric sum we can conclude that 
At step one, create a coreset of size at most $m/2$ of items of weight $2$ and incur an error of at most $c$ or $c \sqrt{m}$.
Then, create a coreset of size $m/4$ of items of weight $4$. Here, you incur error of $2c$ or $2c\sqrt{m/2}  = c\sqrt{2m}$.
Note that the sum of errors is a geometric sequence that is asymptotically dominated by its last element. 
Halting the compression at $\Theta(c/\eps)$ or $\Theta(c^2/\eps^2)$ items respectively achieves the goal.
}
\end{proof}

\begin{fact}
Class discrepancy bounds are tight asymptoticly for {\it unweighted} coreset complexity.
\end{fact}
\begin{proof}Taking for example the bound of Fact~\ref{fct:eps1}, if we can guarantee the existence of an unweighted coreset of size $c/\eps$, then for $m$ items a coreset of size $m/2$ provides a sign assignment with an error of $\eps=2c/m$, leading to an upper bound of $O(c/m)$ on the class discrepancy.
\end{proof}

\add{
The following it a straight forward fact which is provided mainly for convenience. 
It loosely says that optimizing models on coresets generalizes. In other words, ERM works as expected.
\begin{fact}
Let $f(x,q)$ be the loss suffered by model $q$ on example $x$. 
Let $R(q) = \frac{1}{n}F(q) = \frac{1}{n} \sum_{i=1}^n f(x_i,q)$ be the {\it empirical} risk associated with it.
Let $q^*$ denote the best empirical risk minimizer on the data ($q^* = \arg\min_q F(q)$).
Let $\tilde q$ be the minimizer of $q$ over an optimal weighted coreset of size $m$ ($\tilde q = \arg\min_q \tilde F(q)$).
Then $R(\tilde q) \le R(q^*) + O(D_m)$.
\end{fact}
\begin{proof}
This fact follows from the standard argument about empirical risk minimization.
\[
R(\tilde q) = \frac1n F(\tilde q)  \le  \frac{1}{n} \tilde F(\tilde q) + O(D_m) \le \frac{1}{n} \tilde F(q^*) + O(D_m) \le  \frac{1}{n}  F(q^*) + O(D_m)  = R(q^*)+ O(D_m)
\]
The first and last transitions are by definition. The second and fourth are by the approximation bounds above.
The third transition is due to the optimality of $\tilde q$ for $\tilde F$
\end{proof}
}
